---
title: "BTS 510 Lab 6"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* Describe the logic of **linear regression** 
* Describe the **assumptions** of linear regression
* Briefly **present results** of linear regression
* Use a **regression equation** to summarize a model
* Compare and contrast **observed**, **predicted**, and **residual** values

## Data 

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

## Tasks

1. **Does blood pressure differ between those with a suspected infection and those without?** *Fit a linear regression to answer this question.*

```{r}
library(Stat2Data)
data(ICU)
m1 <- lm(SysBP ~ Infection, data = ICU)
summary(m1)
```

* Confidence intervals

```{r}
confint(m1)
```


2. Write up the **interpretation** of the model, including:
  * Intercept value, test statistic, $p$-value
  * Intercept interpretation (in words)
  * Slope value, test statistic, $p$-value
  * Slope interpretation (in words)
  * $R^2$ value, test statistic, $p$-value
  * $R^2$ interpretation (in words)
  
* Intercept ($b_0$) = 138.672, t(198) = 46.440, p<.0001
  * The expected blood pressure for a patient **without suspected infection** (i.e., `Infection` = 0) is 138.672
* Slope ($b_1$) = -15.220, t(198) = -3.303, p = .001134
  * For each 1 unit increase in `Infection`, the expected blood pressure **decreases** by 15.220 points
  * Patients **with suspected infection** (i.e., `Infection` = 1) have blood pressure that is 15.220 points **lower** than patients **without suspected infection** (i.e., `Infection` = 0)
* $R^2$ = 0.05, F(1, 198) = 10.91, p = .001134
  * This model accounts for about 5% of the variance in blood pressure
  * This is significant (meaning that it is different from 0)

3. What is the **predicted blood pressure** for someone *with a suspected infection*? What is the **predicted blood pressure** for someone *without a suspected infection*?

$$\hat{SysBP} = 138.672 - 15.220 \times Infection$$

```{r}
infect_bp <- 138.672 - 15.220 * 1
noinfect_bp <- 138.672 - 15.220 * 0
infect_bp
noinfect_bp
```


4. Calculate the **predicted** and **residual** values for each person and add them to the original dataset. **Does it looks like the assumptions are satisfied?** Make plots to help you decide:
  * Residual vs predictor
  * Q-Q plot of residuals
  * Histogram of residuals (all together and separately for the two groups)

```{r}
ICU$pred <- fitted(m1)
ICU$resi <- residuals(m1)
head(ICU)
```

* Residual vs predictor: Want flat line, grey band even across values of X

```{r}
ggplot(data = ICU, 
       aes(x = Infection, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

* Q-Q plot: Want dots to fall on line

```{r}
ggplot(data = ICU,
       aes(sample = resi)) + 
       stat_qq() +
       stat_qq_line()
```

* Histogram: Want roughly normal (bell-shaped) distributions

```{r}
ggplot(data = ICU,
       aes(x = resi)) +
  geom_histogram()
```

```{r}
ggplot(data = ICU,
       aes(x = resi)) +
  geom_histogram() +
  facet_grid(cols = vars(Infection))
```


5. **Does blood pressure differ between those with a suspected infection and those without? How?** (Brief, simple words, no jargon, no statistics.) 

Yes, blood pressure is different for patients with a suspected infection versus patients without a suspected infection. Patients **with a suspected infection** have **lower** blood pressure levels (on average) than patients **without a suspected infection**.

**Bonus**. A linear regression with only 1 binary predictor (e.g., `Infection`) is **exactly equivalent** to a (independent samples) $t$-test. (Notice that the test for the coefficient / slope in the regression **is** a $t$-test -- this is not a coincidence.) You just get slightly different pieces of information from the two analyses. Run the $t$-test for this same analysis to see what is the same and what is different. This uses the `t.test()` function from the **stats** package (always loaded by default).

`t.test(ICU$SysBP ~ ICU$Infection, alternative = "two.sided", var.equal = TRUE)`

```{r}
t.test(ICU$SysBP ~ ICU$Infection, alternative = "two.sided", var.equal = TRUE)
```

