---
title: "BTS 510 Lab 9"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Describe** *maximum likelihood estimation* for linear regression 
* **Describe** *hypothesis testing* for linear regression coefficients, including the sampling distribution used and degrees of freedom

## Data 

* `FirstYearGPA` data from the **Stat2Data** package: $n$ = 219 subjects
  * `GPA`: First-year college GPA on a 0.0 to 4.0 scale
  * `HSGPA`: High school GPA on a 0.0 to 4.0 scale
  * `SATV`: Verbal/critical reading SAT score
  * `SATM`: Math SAT score
  * `Male`: 1= male, 0= female
  * `HU`: Number of credit hours earned in humanities courses in high school
  * `SS`: Number of credit hours earned in social science courses in high school
  * `FirstGen`: 1= student is the first in her or his family to attend college, 0=otherwise
  * `White`: 1= white students, 0= others
  * `CollegeBound`: 1=attended a high school where >=50% students intended to go on to college, 0=otherwise

## Tasks

Question 1: How do demographic variables (`Male`, `FirstGen` and `White`) predict first year college GPA (`GPA`)?

Question 2: How does HS GPA (`HSGPA`) predict first year college GPA (`GPA`) over demographic variables (`Male`, `FirstGen` and `White`)?

1. Run the two models above.

```{r}
library(Stat2Data)
data(FirstYearGPA)
m1 <- lm(data = FirstYearGPA, GPA ~ Male + FirstGen + White)
summary(m1)
m2 <- lm(data = FirstYearGPA, GPA ~ Male + FirstGen + White + HSGPA)
summary(m2)
```

2. What are the log-likelihoods of each model? What can you say about the models based on those values? What *can't* you say?

```{r}
logLik(m1)
logLik(m2)
```

* We **can say** that model 2 has a (qualitatively) higher (closer to 0) log-likelihood, which is better
  * Without the LRT, we can't say that one is quantitatively higher
  * (We'll do that next)
* We **can't say** that either model is **good** because LL is *relative*
  * You can only use it to *compare* two models

3. Compare the two models using a likelihood ratio test (LRT). Report the results of the test. What can you say about the models based on the test?

```{r}
anova(m1, m2, test = "LRT")
-2 * (logLik(m1) - logLik(m2))
```

* $\chi^2(1) = 54.1902, p<.0001$
  * The models are significantly different from one another
    * Model 2 is a significantly better model than model 1
  * **If they were not significantly different from one another, use the simpler model**

4. Report the results for the better model (based on the LRT). Include all regression coefficients, $R^2$, test statistics, $p$-values.

```{r}
summary(m2)
```

* Based on the likelihood ratio test, model 2 (with 4 predictors) is better than model 1 (with 3 predictors), $\chi^2(1) = 54.1902, p<.0001$
* Sex (`Gender`) is not a significant predictor of first year college GPA, $b_1 = 0.06194, t(214) = 1.144, p>.05$
* First generation status (`FirstGen`) is a significant predictor of first year college GPA, $b_2 = -0.18508, t(214) = -2.125, p<.05$
  * First generation students (`FirstGen` = 1) have **lower** first year college GPA than non-first generation students (`FirstGen` = 0), *controlling for all other predictors*
* Race (`White`) is a significant predictor of first year college GPA, $b_3 = 0.25738, t(214) = 3.786, p<.0001$
  * White students (`White` = 1) have **higher** first year college GPA than non-white students (`White` = 0), *controlling for all other predictors*
* High school GPA is a significant predictor of first year college GPA, $b_4 = 0.55967, t(214) = 7.751, p<.0001$
  * Higher high school GPA is associated with **higher** first year college GPA, *controlling for all other predictors*
* Overall, these four predictors account for 28.84% of the variance in first year college GPA, $R^2 = 0.2884, F(4, 214) = 21.68, p<.0001$
