---
title: "BTS 510 Lab 7"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Add predictors** to a linear regression model
* Interpret **partial regression coefficients**
* **Re-code predictors** to answer your research questions
* **Compare** different models using $R^2_{change}$

## Data 

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

```{r}
library(Stat2Data)
data(ICU)
```

## Research question

**How do infection status (`Infection`) and admission type (`Emergency`) predict blood pressure?**

## Tasks

1. Conduct **three linear regression models** to address the above research questions.

* Model 1: Infection status predicts blood pressure
* Model 2: Admission type predicts blood pressure
* Model 3: Infection status *and* admission type predict blood pressure

```{r}
m1 <- lm(data = ICU, SysBP ~ Infection)
m2 <- lm(data = ICU, SysBP ~ Emergency)
m3 <- lm(data = ICU, SysBP ~ Infection + Emergency)
```


2. **Which is the best model?** 

* Use the $F$-test for $R^2_{change}$ and the AIC values to help you decide. Report the tests and values that you used to decide this. 

```{r}
anova(m1, m2)
anova(m1, m3)
anova(m2, m3)
```

* Note: Cannot use this to compare models with same number of predictors because degrees of freedom is the difference in number of predictors 
  * Comparison of `m1` and `m2` is useless
* Can compare `m3` (two predictors) to each of the 1 predictor models
  * `m3` is better than `m1`, F(1, 197) = 4.6668, p<.05
  * `m3` is better than `m2`, F(1, 197) = 8.5677, p<.01
  * Basically, both predictors are better than either one alone


```{r}
AIC(m1)
AIC(m2)
AIC(m3)
```
* Lower is better for AIC and BIC
  * `m3` has the lowest AIC

3. Report the results of the **best** model, including:
  * Intercept value, test statistic, $p$-value, interpretation (in words)
  * Slope value(s), test statistic(s), $p$-value(s), interpretation(s) (in words)
  * $R^2$ value, test statistic, $p$-value, interpretation (in words)

```{r}
summary(m3)
```

* Intercept
  * $b_0$ = 146.194, t(197) = 31.995, p<.0001
  * The predicted blood pressure for a person without an infection (`Infection` = 0) and without an emergency admission (`Emergency` = 0) is 146.194.
* `Infection` effect
  * $b_1$ = -13.553, t(197) = -2.927, p<.01
  * Holding `Emergency` constant, for a 1 unit increase in `Infection` (i.e., going from no infection (0) to infection (1)), we expect the predicted blood pressure to **decrease** by 13.553 points.
* `Emergency` effect
  * $b_2$ = -11.186, t(197) = -2.160, p<.05
  * Holding `Infection` constant, for a 1 unit increase in `Emergency` (i.e., going from regular admission (0) to emergency admission (1)), we expect the predicted blood pressure to **decrease** by 11.186 points.  
* $R^2$ = 0.074, F(2, 197) = 7.89, p<.001
  * The two predictors together account for 7.4% of the variance in blood pressure.


4. Based on **model 3**, what are the predicted blood pressures for the 4 combinations of infection status and admission type? (i.e., infected non-emergency, non-infected non-emergency, infected emergency, non-infected emergency)

* Method 1: Use R like a calculator
  * Easy and straight-forward
  * You have to keep track of things
  * Not in a data frame to do other things with

```{r}
b0 <- 146.194
b1 <- -13.553
b2 <- -11.186
p_00 <- b0 + b1*0 + b2*0
p_10 <- b0 + b1*1 + b2*0
p_01 <- b0 + b1*0 + b2*1
p_11 <- b0 + b1*1 + b2*1

# no infection, no emergency
p_00
# infection, no emergency
p_10
# no infection, emergency
p_01
# infection, emergency
p_11
```

* Method 2: Use the `predict()` function
  * Step 1. Create a new dataset with the predictor values you want
    * Variables must have same names as in the model: `Infection` and `Emergency`
    * Check the new dataset to make sure it looks like what you want
  * Step 2. Use `predict()` function to add a new variable (`pred_BP`) to the new dataset
    * `new_data$pred_BP` is a variable called `pred_BP` in the `new_data` dataset
    * If this variable doesn't exist already (here, it doesn't), it will be created
    
```{r}
# step 1
new_data <- data.frame(Infection = c(0, 1, 0, 1), Emergency = c(0, 0, 1, 1))
new_data
```

```{r}
# step 2
new_data$pred_BP <- predict(m3, newdata = new_data)
new_data
```

* Use this dataset to make a plot
  * And add some labels

```{r}
ggplot(data = new_data, aes(x = Infection, 
                            y = pred_BP, 
                            shape = as.factor(Emergency))) +
  geom_point(size = 4) +
  geom_line() +
  geom_text(aes(label = round(pred_BP, 2)), 
            nudge_x = 0.25,
            nudge_y = 2) +
  xlim(-0.5, 1.5) +
  ylim(120, 160) +
  labs(shape = "Emergency", 
       y = "Predicted systolic blood pressure (mm/Hg)")
```

* Effect of `Infection`
  * $b_1$ = `r round(b1, 2)`
  * Effect of `Infection`, controlling for `Emergency` is the **slope** of the lines
    * When `Emergency` = 0: 132.64 - 146.19 = `r round(b1,2)`
    * When `Emergency` = 1: 121.45 - 135.01 = `r round(b1,2)`
* Effect of `Emergency`
  * $b_2$ = `r round(b2, 2)`
  * Effect of `Emergency`, controlling for `Infection` is the **difference** between lines
    * When `Infection` = 0: 135.01 - 146.19 = `r round(b2,2)`
    * When `Infection` = 1: 121.45 - 132.64 = `r round(b2,2)`

5. Using the **predicted** and **residual** values, assess whether the assumptions are met for **model 3**. Include the plots and describe your conclusions based on them. 

```{r}
ICU$pred1 <- fitted(m3)
ICU$resi1 <- residuals(m3)
```

::: {.callout-note}
* With **caution**, you can use the `warning: false` option to suppress all the warnings that the loess line (`geom_smooth()`) is giving
* `#| warning: false` in the first line of the code chunk
  * Note the `#|` at the start of the line
  * Unfortunately doesn't show in output, so look at **code**
  * I did this for these first couple of residuals plots because they give warnings because they are trying to calculate a running average at the edges 
:::

* Residuals vs predicted value ($\hat{Y}$)
  * Want a flat line with roughly equal variance (vertical spread) across 

```{r}
#| warning: false
ggplot(data = ICU, 
       aes(x = pred1, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

* Residuals vs `Infection`
  * Want a flat line with roughly equal variance (vertical spread) across 

```{r}
#| warning: false
ggplot(data = ICU, 
       aes(x = Infection, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

* Residuals vs `Emergency`
  * Want a flat line with roughly equal variance (vertical spread) across 

```{r}
#| warning: false
ggplot(data = ICU, 
       aes(x = Emergency, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

* Histogram of residuals, conditional on `Infection`
  * Want roughly normal distributions for each

```{r}
ggplot(data = ICU,
       aes(x = resi1)) +
  geom_histogram() +
  facet_grid(cols = vars(Infection))
```

* Histogram of residuals, conditional on `Emergency`
  * Want roughly normal distributions for each

```{r}
ggplot(data = ICU,
       aes(x = resi1)) +
  geom_histogram() +
  facet_grid(cols = vars(Emergency))
```

6. In plain language, answer the research question: **How do infection status and admission type predict blood pressure?** (This should not be lengthy -- just a couple sentences.)

* Whether someone has an infection or not affects blood pressure
  * Patients *with an infection* had **lower** blood pressure than patients *without an infection*
* Whether someone was an emergency admission or not affects blood pressure
  * Patients with *emergency* admissions had **lower** blood pressure than patients with *non-emergency* admissions
* (Note that this model only includes simple **additive** effects, which leads to the parallel lines. What if it instead looked like, e.g., emergency admission doesn't affect blood pressure for people with infections, but it does for people without infections? **Non-parallel lines**)
